---
title: "Build and deploy a stroke prediction model using R"
date: "`r Sys.Date()`"
output: html_document
author: "Shifa Maqsood"
---

# About Data Analysis Report

This RMarkdown file contains the report of the data analysis done for the project on building and deploying a stroke prediction model in R. It contains analysis such as data exploration, summary statistics and building the prediction models. The final report was completed on `r date()`. 

**Data Description:**

According to the World Health Organization (WHO) stroke is the 2nd leading cause of death globally, responsible for approximately 11% of total deaths.

This data set is used to predict whether a patient is likely to get stroke based on the input parameters like gender, age, various diseases, and smoking status. Each row in the data provides relevant information about the patient.



## Load data and install packages

```{r}
df_strokemodel <- read.csv("healthcare-dataset-stroke-data.csv")
library(dplyr)
library(tidyverse)
library(ggplot2)
library(caret)
library(reshape2)
library(randomForest)
library(gridExtra)


```


```{r}
df_strokemodel <- read.csv("healthcare-dataset-stroke-data.csv")
library(dplyr)
```


## Pre-Processing 

```{r}
#drop Id
df_strokemodel <- df_strokemodel %>%
  select(-id)

#drop other 
df_strokemodel <- df_strokemodel[!df_strokemodel$gender == 'Other',]


df_strokemodel$bmi <- as.numeric(df_strokemodel$bmi)

# converting "N/A" to "NA"
df_strokemodel$bmi[df_strokemodel$bmi == "N/A"] <-  NA

#replace bmi NA values with median values
bmi_median <-  median(df_strokemodel$bmi, na.rm = TRUE)
df_strokemodel$bmi[is.na(df_strokemodel$bmi)] <- bmi_median

# converting variables stroke,hypertension and heart disease to categorical factors 
df_strokemodel$stroke <- ifelse(df_strokemodel$stroke == 1, 'stroke', 'no stroke') 
df_strokemodel$hypertension <- ifelse(df_strokemodel$hypertension == 1, 'hypertension', 'no hypertension') 
df_strokemodel$heart_disease <- ifelse(df_strokemodel$heart_disease == 1, 'heart disease', 'no heart disease') 

# convert from  numeric/character to factor.  check the type  by using str()

df_strokemodel$bmi <- as.numeric(df_strokemodel$bmi)

df_strokemodel$gender <- as.factor(df_strokemodel$gender )
df_strokemodel$ever_married <- as.factor(df_strokemodel$ever_married)
df_strokemodel$work_type <- as.factor(df_strokemodel$work_type)
df_strokemodel$Residence_type <- as.factor(df_strokemodel$Residence_type)
df_strokemodel$smoking_status <- as.factor(df_strokemodel$smoking_status)

```

## Exploratory Data Analysis 
```{r}
#Conducting Data Visualization helps visualize the basic visual information. 

graph1 <- ggplot(df_strokemodel, aes(x=" ", y = gender, fill= gender)) + geom_bar(stat="identity", width=1)  + coord_polar("y", start=0)

graph2 <- ggplot(df_strokemodel, aes(x="", y =hypertension, fill= hypertension))+ geom_bar(stat="identity", width=1)  + coord_polar("y", start=0)



graph3 <- ggplot(df_strokemodel, aes(x="", y =heart_disease, fill= heart_disease))+ geom_bar(stat="identity", width=1)  + coord_polar("y", start=0)


graph4 <- ggplot(df_strokemodel, aes(x="", y =ever_married, fill= ever_married))+ geom_bar(stat="identity", width=1)  + coord_polar("y", start=0)

graph5 <- ggplot(df_strokemodel, aes(x="", y =Residence_type, fill= Residence_type))+ geom_bar(stat="identity", width=1)  + coord_polar("y", start=0)

graph6 <- ggplot(df_strokemodel, aes(x="", y =stroke, fill= Residence_type))+ geom_bar(stat="identity", width=1)  + coord_polar("y", start=0)



grid.arrange(graph1,graph2,graph3, ncol = 3)
grid.arrange(graph4, graph5, graph6, ncol=3)

```

```{r}
#exploring relation between stroke and other variables 

graph_1 <- ggplot(df_strokemodel)+ geom_bar(mapping = aes(x= gender, fill= stroke))
graph_2 <- ggplot(df_strokemodel)+ geom_bar(mapping = aes(x= smoking_status, fill= stroke))
graph_3 <- ggplot(df_strokemodel)+ geom_bar(mapping = aes(x= heart_disease, fill= stroke))
graph_4 <- ggplot(df_strokemodel)+ geom_bar(mapping = aes(x= hypertension, fill= stroke))
graph_5 <- ggplot(df_strokemodel)+ geom_bar(mapping = aes(x= work_type, fill= stroke))
graph_6 <-ggplot(df_strokemodel)+ geom_bar(mapping = aes(x= Residence_type, fill= stroke))



grid.arrange(graph_1, graph_2, graph_3, ncol=2)

grid.arrange(graph_4, graph_5, graph_6, ncol= 2)



age_graph <- ggplot(data = df_strokemodel) + geom_histogram(mapping = aes(x = age), binwidth = 0.5)

```

## Pearson's Correlation 
```{r}
#computing Pearson's correlation between all the variables 

#first convert categorical to  character  variable 
copyofdf_strokemodel <- df_strokemodel



suppressWarnings(copyofdf_strokemodel$gender <- (as.character(copyofdf_strokemodel$gender)))

suppressWarnings(copyofdf_strokemodel$ever_married <- (as.character(copyofdf_strokemodel$ever_married)))

suppressWarnings(copyofdf_strokemodel$Residence_type <- (as.character(copyofdf_strokemodel$Residence_type)))

suppressWarnings(copyofdf_strokemodel$work_type <- (as.character(copyofdf_strokemodel$work_type)))

suppressWarnings(copyofdf_strokemodel$smoking_status <- (as.character(copyofdf_strokemodel$smoking_status)))

#then convert to binary then convert to numeric 
  
  #gender
copyofdf_strokemodel$gender[copyofdf_strokemodel$gender =="Male"] <-0
copyofdf_strokemodel$gender[copyofdf_strokemodel$gender =="Female"] <-1

copyofdf_strokemodel$gender <- as.numeric(copyofdf_strokemodel$gender)

  #hypertension 
copyofdf_strokemodel$hypertension[copyofdf_strokemodel$hypertension == "no hypertension"] <- 0
copyofdf_strokemodel$hypertension[copyofdf_strokemodel$hypertension == "hypertension"] <-1

copyofdf_strokemodel$hypertension <- as.numeric(copyofdf_strokemodel$hypertension)

  #heart_disease 

copyofdf_strokemodel$heart_disease[copyofdf_strokemodel$heart_disease == "no heart disease"] <- 0
copyofdf_strokemodel$heart_disease[copyofdf_strokemodel$heart_disease == "heart disease"] <-1

copyofdf_strokemodel$heart_disease <- as.numeric(copyofdf_strokemodel$heart_disease)


  #stroke 

copyofdf_strokemodel$stroke[copyofdf_strokemodel$stroke == "no stroke"] <- 0
copyofdf_strokemodel$stroke[copyofdf_strokemodel$stroke == "stroke"] <-1

copyofdf_strokemodel$stroke <-as.numeric(copyofdf_strokemodel$stroke)

  #residence 


copyofdf_strokemodel$Residence_type[copyofdf_strokemodel$Residence_type == "Urban"] <- 0
copyofdf_strokemodel$Residence_type[copyofdf_strokemodel$Residence_type == "Rural"] <-1

copyofdf_strokemodel$Residence_type <-as.numeric(copyofdf_strokemodel$Residence_type)

  #ever_married

copyofdf_strokemodel$ever_married[copyofdf_strokemodel$ever_married == "No"] <- 0
copyofdf_strokemodel$ever_married[copyofdf_strokemodel$ever_married == "Yes"] <-1

# turn them into numeric after coverting  into factor or else values will be replaced by NA

copyofdf_strokemodel$ever_married <-as.numeric(copyofdf_strokemodel$ever_married)


#change unknown to never smoked 

copyofdf_strokemodel <- copyofdf_strokemodel %>%
  mutate(smoking_status = replace(smoking_status, smoking_status == "Unknown", "never smoked"))
 
# drop work type and smoking status 
copyofdf_strokemodel_quantvar <- subset(copyofdf_strokemodel, select = -c(work_type, smoking_status))


copyofdf_strokemodel_cor <- round(cor(copyofdf_strokemodel_quantvar),2)

library(ggplot2)

#upload libraries again for the corr heat map

stroke_corr_heatmap <- ggplot(data=reshape2::melt(copyofdf_strokemodel_cor),aes(x=Var1, y=Var2, fill=value)) + geom_tile() +  scale_fill_gradient2(low = "#547a0a", high = "#bc7543", mid = "white", midpoint = 0, limit = c(-1,1), space = "Lab", name="Pearson\nCorrelation") + geom_text(aes(Var2, Var1, label = value), color = "black", size = 4) + theme(axis.text.x = element_text(angle = 30))

stroke_corr_heatmap

#The correlation heatmap shows us that there is a correlation between age and ever married which is expected however the value of 0.25 for stroke and age shows a hint of correlation. Whereas we see residence type and stroke have a correlation of -0.02. 


```



```{r}
# Are people with high bmi and age  likely to get a stroke?

bmi_age_plot <- ggplot(data = df_strokemodel, aes(x = age, y= bmi, color = stroke))+geom_point()

bmi_age_plot

#this plot clearly shows that as patients age, they have a high chance of getting a stroke,
#although with bmi, there is a also a likelihood of getting a stroke with a lesser bmi but chances increase as the age and bmi increase.

```


## Logistic Regression  for Modelling 

```{r}
##pre-processing for Logistic Regression
#make more dummy variables since we can only use binary 
# for work type and smoking status 

dummydataset <- copyofdf_strokemodel %>%
      mutate(work_type = case_when(work_type == "children" ~0,
                               work_type == "Never_worked" ~1,
                               work_type == "Private" ~2,
                               work_type == "Govt_job" ~3,
                               work_type == "Self-employed" ~4),
             smoking_status = case_when(smoking_status == "never smoked" ~0,
                                    smoking_status == "formerly smoked" ~1,
                                    smoking_status == "smokes" ~2,
                                    smoking_status == "Unknown" ~3))
data_regression_model <- dummydataset[,c("age","hypertension","avg_glucose_level","heart_disease","smoking_status", "stroke")]

## data_regression_model includes the variables we need for our regression 


```

```{r}

Training <- createDataPartition(y = data_regression_model$stroke, p= 0.7, list = FALSE)

training <- data_regression_model[Training,]
testing <- data_regression_model[-Training,]

#now we have two variables, testing and training for our regression model
#we will train it on training data set
#training is 70% of our original  dataset as dim(training) can be used to verify 

model <- glm(stroke~., family = binomial(link = 'logit'), data = training)

summary(model)



```



```{r}
anova(model, test = "Chisq")
```




```{r}
model.prob <- predict(model, testing, type = "response")
```


```{r}
confusionMatrix(data = as.factor(as.numeric(model.prob>0.5)), reference = as.factor(testing$stroke))
```


## Random Foresst for Modelling

```{r}


##Alternatively, we can use random forest too 

#change into factors 
df_strokemodel_randomForest <- df_strokemodel
df_strokemodel_randomForest$hypertension <- as.factor(df_strokemodel_randomForest
                                                      $hypertension)

df_strokemodel_randomForest$heart_disease <- as.factor(df_strokemodel_randomForest
                                                      $heart_disease)

df_strokemodel_randomForest$stroke <- as.factor(df_strokemodel_randomForest
                                                      $stroke)

n_obs <- nrow(df_strokemodel_randomForest)
split <- round(n_obs*0.7)
train <- df_strokemodel_randomForest[1:split,]

test <- df_strokemodel_randomForest[(split+1):nrow(df_strokemodel_randomForest),]

dim(train)
dim(test)





```

```{r}

set.seed(123)

rf_model<-randomForest(formula= stroke~.,data = train, na.action=na.roughfix)
rf_model
```

## Conufusion Matrix 

```{r}
confusionMatrix(predict(rf_model, test), test$stroke)
```
















